---
title: "Careers over time: transitions, promotions, and churn"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Careers over time: transitions, promotions, and churn}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(synthid)
```

Once [`link_panel()`](../reference/link_panel.html) has stamped a stable `EMP_ID`
on every record, the panel stops being a pile of person-years and becomes a set of
*careers* you can read over time. Three functions answer three questions:

- **[`flag_position_transitions()`](../reference/flag_position_transitions.html)** —
  when does the *person* holding a seat change? (succession)
- **[`flag_promotions()`](../reference/flag_promotions.html)** — when does a
  *person's own title* genuinely advance? (promotion, net of labeling noise)
- **[`position_churn()`](../reference/position_churn.html)** — how does a *body*
  such as a board turn over year to year? (churn)

All three read change off the `EMP_ID` clusters, so they run on the output of
`link_panel()`. The examples below hand-stamp `EMP_ID` to stay deterministic; in
practice it comes from the linkage step.

## A small linked panel

One record per organization-year per person, exactly as `link_panel()` returns.
Organization `A` has an Executive Director seat and a five-person board.

```{r panel}
rec <- function(ein, taxyr, EMP_ID, name, title.standard) {
  data.frame(ein, taxyr, EMP_ID, name, title.standard, stringsAsFactors = FALSE)
}
linked <- rbind(
  # --- the Executive Director seat: ALICE, then a vacant year, then BOB ---
  rec("A", 2018, "E-ALICE", "ALICE REYES", "EXECUTIVE DIRECTOR"),
  rec("A", 2019, "E-ALICE", "ALICE REYES", "EXECUTIVE DIRECTOR"),
  rec("A", 2021, "E-BOB",   "BOB NGUYEN",  "EXECUTIVE DIRECTOR"),
  # --- the board, with some turnover ---
  rec("A", 2018, "E-P", "PAT LEE",   "TRUSTEE"),
  rec("A", 2018, "E-Q", "QUINN ABE", "TRUSTEE"),
  rec("A", 2018, "E-R", "RILEY COX", "BOARD PRESIDENT"),
  rec("A", 2018, "E-S", "SAM DIAZ",  "TRUSTEE"),
  rec("A", 2018, "E-T", "TONI FALK", "TRUSTEE"),
  rec("A", 2019, "E-P", "PAT LEE",   "TRUSTEE"),
  rec("A", 2019, "E-Q", "QUINN ABE", "TRUSTEE"),
  rec("A", 2019, "E-R", "RILEY COX", "BOARD PRESIDENT"),
  rec("A", 2019, "E-S", "SAM DIAZ",  "TRUSTEE"),
  rec("A", 2019, "E-U", "UMA GHOSH", "TRUSTEE"),      # U in, T out
  rec("A", 2020, "E-P", "PAT LEE",   "TRUSTEE"),
  rec("A", 2020, "E-Q", "QUINN ABE", "TRUSTEE"),
  rec("A", 2020, "E-R", "RILEY COX", "BOARD PRESIDENT"),
  rec("A", 2020, "E-V", "VIC HALE",  "TRUSTEE"),      # V, W in; S, U out
  rec("A", 2020, "E-W", "WES IRWIN", "TRUSTEE"),
  rec("A", 2021, "E-P", "PAT LEE",   "TRUSTEE"),
  rec("A", 2021, "E-Q", "QUINN ABE", "TRUSTEE"),
  rec("A", 2021, "E-R", "RILEY COX", "BOARD PRESIDENT"),
  rec("A", 2021, "E-V", "VIC HALE",  "TRUSTEE"),
  rec("A", 2021, "E-W", "WES IRWIN", "TRUSTEE")
)
```

## 1. Position transitions

A *transition* is a **new `EMP_ID` taking a seat that a different person held
before** — the title stays put, the person underneath it changes. Point the
function at the seat by title (or by `role`, e.g. `role = "BOARD"`):

```{r transitions}
tr <- flag_position_transitions(linked, position = "EXECUTIVE DIRECTOR")
tr[, c("taxyr", "EMP_ID", "prev_year", "incoming", "transition", "predecessor")]
```

`ALICE` in 2018 is `incoming` (first occupant) but **not** a transition — there was
no predecessor to succeed. The interesting case is the vacancy: the seat is empty
in 2020, then `BOB` appears in 2021. By default the reference point is the prior
**filing** year (2020, when the seat was vacant), so `BOB` reads as a fresh
establishment, not a succession:

```{r transitions-strict}
subset(tr, EMP_ID == "E-BOB", c(taxyr, prev_year, transition, predecessor))
```

Set `gap_tolerant = TRUE` to measure against the last year the seat was actually
**occupied**, bridging the vacancy back to `ALICE`:

```{r transitions-gap}
trg <- flag_position_transitions(linked, position = "EXECUTIVE DIRECTOR",
                                 gap_tolerant = TRUE)
subset(trg, EMP_ID == "E-BOB",
       c(taxyr, prev_year, years_since_prev, transition, predecessor))
```

Now `BOB`'s arrival is a transition, its `predecessor` is `ALICE`, and
`years_since_prev` records the two-year span it bridged. Use strict when a vacancy
should *not* read as a handover; use gap-tolerant to follow a seat across gaps.

Transitions work best on a one- or two-holder seat. For a whole body, measure
churn instead.

## 2. Position churn

`position_churn()` accounts for turnover in a *set* of people. Select the board by
its role and it reports, per year, how many hold the seat, and — against the prior
year — how many are **new**, how many **left**, and how many are **stable**:

```{r churn}
position_churn(linked, role = "BOARD")
```

Read 2019: the board still has five members, but one is new (`UMA`), one left
(`TONI`), and four are stable — `retention = 4/5`. 2020 is a bigger shake-up
(`SAM` and `UMA` out, `VIC` and `WES` in); 2021 is fully stable. The first year is
`NA` for the flow columns because there is no prior year to compare against.

Pass `by_org = FALSE` to pool the whole panel into one population-level churn
series instead of one per organization (pass a cross-org id to `id=` if you want to
avoid counting a person once per organization).

## 3. Promotions

A person's title changing is *not* enough to call a promotion: titles are reported
inconsistently year to year, so a naive diff over-counts wildly.
`flag_promotions()` inspects each person's **upstream** history and compares
**both** the raw and standardized titles to sort every change into one of four
kinds. It needs the richer columns (`title.raw`, `tot.comp`):

```{r promo-data}
prec <- function(ein, taxyr, EMP_ID, std, raw = std, comp = NA_real_) {
  data.frame(ein, taxyr, EMP_ID, name = EMP_ID,
             title.standard = std, title.raw = raw, tot.comp = comp,
             stringsAsFactors = FALSE)
}
careers <- rbind(
  # a real promotion: paid role move with a pay jump
  prec("B", 2018, "E-X", "PROGRAM DIRECTOR",        comp = 60000),
  prec("B", 2019, "E-X", "EXECUTIVE DIRECTOR",      comp = 95000),
  prec("B", 2020, "E-X", "EXECUTIVE DIRECTOR",      comp = 96000),
  # a standardizer relabel: raw title identical, standard label wobbled
  prec("C", 2018, "E-Y", "CHIEF EXECUTIVE OFFICER", raw = "CEO", comp = 1e5),
  prec("C", 2019, "E-Y", "EXECUTIVE DIRECTOR",      raw = "CEO", comp = 1e5),
  # noise: a one-year blip that reverts, and a label seen upstream
  prec("D", 2018, "E-Z", "MANAGER"),
  prec("D", 2019, "E-Z", "DIRECTOR OF OPERATIONS"),
  prec("D", 2020, "E-Z", "MANAGER")
)
pr <- flag_promotions(careers)
pr[, c("EMP_ID", "from_year", "from_title", "to_title",
       "change_type", "direction", "comp_change")]
```

The `change_type` factor separates the signal from three kinds of noise:

- **`real`** — a novel, persistent title move. `E-X` going PROGRAM DIRECTOR →
  EXECUTIVE DIRECTOR with pay $60k → $95k is a `promotion` (direction is read from
  the compensation change when present, else a coarse role rank).
- **`relabel`** — `E-Y`'s CEO → Executive Director is the *same* post: the raw
  title (`CEO`) never changed, only the standardizer's label did. Agreement on
  **either** the raw or standardized view marks it a relabel, not a move.
- **`transient` / `oscillation`** — `E-Z`'s MANAGER → DIRECTOR OF OPERATIONS
  reverts the next year (`transient`), and the return to MANAGER matches a label
  held upstream (`oscillation`). Both are flip-flopping labels, not advancement.

Filter to what you mean by a promotion:

```{r promo-real}
subset(pr, change_type == "real" & direction == "promotion",
       c(EMP_ID, from_title, to_title, comp_change))
```

## Reading the print summaries

Each function returns a classed data frame whose `print` method leads with a
compact summary before the rows — handy at the console:

```{r print}
flag_promotions(careers)
```

## Sanity-checking the flag against pay

A useful independent check on the transition flag: in a nonprofit the chief
executive is *usually* the highest-paid person, so following the **top earner** as
its own synthetic seat should broadly agree with following the CEO/ED title. You
can build that seat with the same machinery — mark the highest-paid `EMP_ID` per
organization-year and run `flag_position_transitions()` on it:

```{r topearner}
top_earner_seat <- function(df) {
  paid <- df[!is.na(df$tot.comp) & df$tot.comp > 0, ]
  by <- paste(paid$ein, paid$taxyr, sep = "\r")
  top <- do.call(rbind, lapply(split(paid, by), function(g) {
    g <- g[order(-g$tot.comp, g$EMP_ID), ][1, ]
    g$title.standard <- "TOP_EARNER"; g
  }))
  flag_position_transitions(top, position = "TOP_EARNER")
}
```

On the full 15-year reference panel the two views agree on ~88–90% of
organization-year steps, and the CEO/ED is the top earner in ~88% of
organization-years. The disagreements are all explainable rather than flag errors:
a highly paid non-executive (a physician, a coach) can out-earn the director, and
in the transition year itself an incoming director is often only partially paid —
so "highest paid" is a noisier seat than the title, which is exactly why the
title-based flag is the primary signal and pay is the corroborating one.

## Practical notes

- **Denominator.** Turnover is only defined for organizations that *have* the seat.
  Only ~1 in 3 filers report a paid CEO/ED at all — the rest are all-volunteer, or
  led by a paid President that [`classify_title_role()`](../reference/classify_title_role.html)
  treats as `BOARD`. Report transitions against organizations-with-a-seat, not all
  organizations.
- **Link quality bounds turnover.** A year the linker fails to connect a person
  reads as one departure plus one arrival, so churn is a slight *over*-estimate of
  true turnover. A 100% single-year board turnover is almost always a linkage or
  filing-gap artifact, not a board replacing itself overnight — inspect those
  rather than trusting them.

## See also

- [`link_panel()`](../reference/link_panel.html) — produce the `EMP_ID` these
  functions read.
- [Title roles](title-roles.html) — the `BOARD` / `OFFICER` / `STAFF` / `OTHER`
  taxonomy used by `role =` selection.
