---
title: "Cross-organization linkage (interlocking directorates)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cross-organization linkage (interlocking directorates)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(synthid)
```

`link_panel()` links a person to themselves *across years within one
organization*. This article covers the complementary problem: linking the same
person *across different organizations* — the basis of **interlocking directorate**
studies, where one individual sits on several nonprofit boards.

The two are deliberately separate. Within-org linkage can lean on a tight block
(same `ein`) and rich cross-year signal. Cross-org linkage cannot — it must search
the whole population — so it uses a different machine: **person profiles** plus
**hash-lookup blocking**.

## The shape of the pipeline

```
link_panel()            -> EMP_ID (within-org, across years)
  |
build_person_profile()  -> one identity profile per EMP_ID, with blocking keys
  |
blocking_passes()       -> hash keys that bring candidate pairs together
  |
score_candidate_pairs() -> score cross-org pairs
  |
filter_edges_by_role()  -> keep the interlocks you care about (e.g. board-board)
  |
stamp_xorg()            -> a cross-org person id
```

`link_cross_org()` wraps the whole chain; the pieces are exported for custom
pipelines.

## Person profiles

A person may have many within-org records. `build_person_profile()` reduces each
`EMP_ID` cluster to one canonical profile — the identity you actually match on
across orgs — and precomputes the blocking keys used to find candidates:

```{r profile, eval = FALSE}
linked   <- link_panel(panel)              # must already carry EMP_ID
profiles <- build_person_profile(linked)
# each profile carries last-name blocking keys for the hash join:
str(profiles$last_name_keys[[1]])
```

## Blocking passes

Comparing every profile to every other is quadratic and hopeless at population
scale. Instead, candidates are generated by **blocking passes** — each pass emits
a hash key, and only profiles sharing a key are ever compared. The available
passes are inspectable:

```{r passes}
names(blocking_passes())
```

Multiple passes give multiple chances to co-locate a true pair (e.g. a
last-name-based key and a phonetic key), so a single noisy field does not hide an
interlock. `person_blocking_keys()` computes the keys for a profile;
`candidate_pairs()` turns shared keys into the pairs to score.

## Scoring and role filtering

Cross-org pairs are scored with the same Fellegi–Sunter machinery as within-org
linkage (see [How the matching works](matching-methodology.html)), using the
population-level surname prior (`population_surname_weight()`) rather than a
per-org one. You then keep the edges whose **role composition** matches your
question — most interlock studies want board-to-board ties:

```{r roles, eval = FALSE}
xo <- link_cross_org(linked, state = "state", ntee = "ntee", threshold = 8)

subset(xo$profiles, XORG_N_ORGS > 1)                 # people who interlock
board_only_edges(xo$edges)                           # board <-> board ties only
non_board_edges(xo$edges)                            # staff/officer ties
```

`filter_edges_by_role()` is the general form; `board_only_edges()` and
`non_board_edges()` are convenience wrappers. Role is assigned by the taxonomy in
[Title roles](title-roles.html).

## Stamping the cross-org id

`stamp_xorg()` closes the accepted cross-org edges into components and stamps a
cross-organization person identifier, alongside a count of the distinct
organizations each person appears in (`XORG_N_ORGS`) — the raw material for an
interlock network.

## See also

- `link_cross_org()`, `build_person_profile()`, `blocking_passes()`,
  `person_blocking_keys()`, `candidate_pairs()`, `score_candidate_pairs()`,
  `filter_edges_by_role()`, `board_only_edges()`, `non_board_edges()`,
  `stamp_xorg()` in the reference.
- [Title roles](title-roles.html) for how BOARD / OFFICER / STAFF is decided.
```
