---
title: "Getting started: a panel walk-through"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started: a panel walk-through}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(synthid)
```

## What `synthid` does

The upstream packages (`titleclassifier`, `peopleparser`) answer *"what is this
record?"* — a parsed name, a standardized title. `synthid` answers a different
question: *"**who** is this record, across years?"*

Given a stacked, multi-year panel of Form 990 director/officer/trustee
compensation rows, it assigns every record a **stable per-person identifier**,
`EMP_ID`, that is consistent across tax years. That identifier is the keystone a
longitudinal panel design rests on: the raw `person.id` in the source data is
*not* stable across years, so a dedicated linkage layer is required.

This article walks the whole pipeline on a tiny synthetic panel you can paste
and run.

## The input: a compensation panel

`synthid` consumes a data frame with, at minimum, an organization id (`ein`), a
tax year (`taxyr`), and a full `name`. If the parsed name components and a
standardized title are present, they are used to sharpen the match. Here is a
small panel with three organizations-worth of behavior baked in:

```{r panel}
panel <- rbind(
  data.frame(ein = "100", taxyr = 2019, name = "JOHN SMITH SR",
             first_name = "JOHN", last_name = "SMITH", suffix = "SR",
             gender = "M", title.standard = "CEO"),
  data.frame(ein = "100", taxyr = 2019, name = "JOHN SMITH JR",
             first_name = "JOHN", last_name = "SMITH", suffix = "JR",
             gender = "M", title.standard = "TREASURER"),
  data.frame(ein = "100", taxyr = 2019, name = "ANN JONES",
             first_name = "ANN", last_name = "JONES", suffix = "",
             gender = "F", title.standard = "SECRETARY"),
  data.frame(ein = "100", taxyr = 2020, name = "JOHN SMITH SR",
             first_name = "JOHN", last_name = "SMITH", suffix = "SR",
             gender = "M", title.standard = "CEO"),
  data.frame(ein = "100", taxyr = 2020, name = "JOHN SMITH JR",
             first_name = "JOHN", last_name = "SMITH", suffix = "JR",
             gender = "M", title.standard = "TREASURER"),
  data.frame(ein = "100", taxyr = 2021, name = "JOHN SMITH SR",
             first_name = "JOHN", last_name = "SMITH", suffix = "SR",
             gender = "M", title.standard = "CEO"),
  # A different organization with a same-named person -- must NOT cross-link.
  data.frame(ein = "200", taxyr = 2019, name = "JOHN SMITH SR",
             first_name = "JOHN", last_name = "SMITH", suffix = "SR",
             gender = "M", title.standard = "PRESIDENT"),
  data.frame(ein = "200", taxyr = 2020, name = "JOHN SMITH SR",
             first_name = "JOHN", last_name = "SMITH", suffix = "SR",
             gender = "M", title.standard = "PRESIDENT"),
  stringsAsFactors = FALSE
)
panel
```

There are three deliberate challenges here:

- **A father/son pair on the same board** (`JOHN SMITH SR` and `JOHN SMITH JR`)
  who share a first *and* last name and differ only by suffix. They must stay
  **two** people.
- **A person observed across three years** (`JOHN SMITH SR`, org 100) who should
  collapse to **one** id spanning 2019-2021.
- **A same-named person at a different organization** (`JOHN SMITH SR`, org 200)
  who must **not** be linked to the org-100 Smiths.

## Link the panel

One call does the whole pipeline — prepare, block, compare, score, match, and
close the accepted links into person clusters:

```{r link}
linked <- link_panel(panel)
linked[, c("ein", "taxyr", "name", "suffix", "EMP_ID", "EMP_N_RECORDS", "EMP_N_YEARS")]
```

Read the added columns:

- **`EMP_ID`** — the stable person identifier (a content hash of the cluster's
  members, so it is reproducible across runs).
- **`EMP_N_RECORDS`** — how many records are in the person's cluster.
- **`EMP_N_YEARS`** — how many distinct tax years the person is observed.

Check that the three challenges resolved correctly:

```{r checks}
# The org-100 SR appears in 3 years under ONE id.
sr100 <- linked$EMP_ID[linked$ein == "100" & linked$suffix == "SR"]
length(unique(sr100)) == 1L

# Father and son (SR vs JR) are DIFFERENT people.
jr100 <- linked$EMP_ID[linked$ein == "100" & linked$suffix == "JR"]
unique(sr100) != unique(jr100)

# The same-named person at org 200 is NOT the org-100 person.
sr200 <- linked$EMP_ID[linked$ein == "200"]
length(intersect(sr100, sr200)) == 0L
```

## The run report

`synthid_report()` prints a summary of what happened — how many clusters formed,
how many span multiple years, and how many links were blocked by the
one-record-per-organization-year guard:

```{r report}
synthid_report(linked)
```

## How it decided (in one paragraph)

Candidate pairs are formed **only within an organization and only across
different tax years** — blocks are small, so every cross-year pair inside an org
is compared. Each field contributes `weight x (2*similarity - 1)`; a missing
field is neutral. Two touches matter for the panel above: the surname weight is
scaled *per organization* by how rare the surname is inside that org, so agreeing
on a common board surname carries little information and the **suffix / first
name / gender** do the discriminating (this is what keeps SR and JR apart); and
because pairs are only formed within an `ein`, the org-200 Smith is never even
compared to the org-100 Smiths. Accepted links are then closed into clusters
under a hard invariant: **a person holds at most one record per
organization-year.**

## A learned alternative: the EM model

The default `method = "weighted"` uses a hand-set additive score. You can instead
fit an **unsupervised** Fellegi-Sunter latent-class model to the candidate
comparisons and link by the learned posterior match probability — no labels
required. It also attaches `EMP_LINK_CONF`, the weakest link probability holding
each cluster together:

```{r em}
linked_em <- link_panel(panel, method = "em")
linked_em$EMP_LINK_CONF

# The learned agreement / disagreement weights:
fs_weights(attr(linked_em, "synthid")$model)
```

The EM fit doubles as a **validation of the hand weights** — on the real
development panel the two methods agree on 99.85% of accept/reject decisions —
and the posterior is sharply bimodal, which is what makes `EMP_LINK_CONF` a
usable confidence signal for a downstream panel model.

## Adapting to your column names

If your panel names the organization or other fields differently, pass a modified
column map. You can also raise the `threshold` for higher precision:

```{r cols}
cols <- synthid_cols()
cols$org_id <- "ein"   # e.g. change to "ein9" for a 9-digit EIN column
str(cols)

# linked <- link_panel(panel, cols = cols, threshold = 8)
```

## Where the identifiers come from

`EMP_ID` answers *who, across years*. There is a second identifier,
`PERSON_YEAR_ID`, that answers *which source row* — a stable back-link from a
linked record to its original Form 990 Part VII line, computed from the source
key at ingest. The two are complementary: `EMP_ID` groups a person's rows
together; `PERSON_YEAR_ID` ties each of those rows back to the raw filing it came
from. See the function reference for
[`person_year_id()`](../reference/person_year_id.html) and
[`create_emp_ids()`](../reference/create_emp_ids.html).

## Where to go next

- The [function reference](../reference/index.html), grouped by subsystem
  (panel linkage, identifiers, match model, name comparison, cross-organization
  linkage, title roles, and link review).
- `link_panel()` for the full set of arguments (`method`, `threshold`,
  `weights`, `prob_threshold`).
- `default_weights()` and `fs_weights()` to inspect and tune the scoring.
```
