---
title: "The two identifiers: EMP_ID and PERSON_YEAR_ID"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The two identifiers: EMP_ID and PERSON_YEAR_ID}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(synthid)
```

`synthid` produces two complementary identifiers. They answer different questions
and should not be confused:

| Identifier | Answers | Grain |
|---|---|---|
| **`PERSON_YEAR_ID`** | *Which source row is this?* | one Form 990 Part VII line |
| **`EMP_ID`** | *Who is this, across years?* | one person (a cluster of rows) |

`PERSON_YEAR_ID` is the stable **back-link** from a linked record to the raw
filing it came from; `EMP_ID` is the stable **cross-year person** id that groups a
person's rows together. Both are deterministic hashes, so they are reproducible
across runs and recoverable from the source if ever lost.

## `PERSON_YEAR_ID`: a back-link to the source row

The key is the native filing/person pair `(OBJECTID, TABLE_ID)`:

- **`OBJECTID`** identifies the filing. It comes from the IRS filing index (not
  the XML body parse), so it is parser-independent.
- **`TABLE_ID`** is the stored per-person key *within* a filing. It is unique
  inside a filing and never empty, so — unlike a name or title key — it needs no
  disambiguation and is unaffected by downstream name/title cleaning.

```{r pyid}
person_year_id(
  object_id = c("OID-201243189349201149", "OID-201243189349201149"),
  table_id  = c("TID-00001",              "TID-00002")
)
```

Two people in the same filing (`TID-00001`, `TID-00002`) get distinct ids. The
recipe is `PYID-` + a hash of `keyspec | OBJECTID | TABLE_ID`; the `keyspec` is
baked in so a future change to the recipe can never silently collide with old ids.

### Why not hash the name?

A name/title key collides whenever two people share a name in the same org-year,
and it *changes* when titles are later cleaned or reclassified. `(OBJECTID,
TABLE_ID)` has neither problem. The raw `OBJECTID` and `TABLE_ID` columns are kept
alongside `PERSON_YEAR_ID` so the id is always auditable and re-derivable.

## Filing resolution: the subtlety `PERSON_YEAR_ID` forced us to fix

`EIN + TAXYR` is *not* a unique filing key. An organization can submit several
returns for one tax year — an original plus amendments — each a **distinct
`OBJECTID` with its own independent `TID-00001..N` sequence**. So the same
`TABLE_ID` number recurs *across* filings, while never repeating *within* one.

Left alone, the same person would appear once per filing, inflating the panel and
letting superseded returns feed stale data downstream. The preprocessing resolves
this by keeping only the **authoritative** filing per org-year — latest
`RETURN_TIME_STAMP`, amended-preferred, with a deterministic final tie-break:

```{r resolve}
# Three filings of the same 2011 return: one original + two amendments.
filings <- data.frame(
  OBJECTID = c("OID-201213189349200431", "OID-201223189349201207", "OID-201243189349201149"),
  amended  = c(FALSE, TRUE, TRUE),
  ts       = c("2012-11-13 14:05:34", "2012-11-13 17:14:29", "2012-11-13 19:16:05"),
  stringsAsFactors = FALSE
)
# Winner rule: latest timestamp, amended-preferred, then max OBJECTID.
o <- order(filings$ts, filings$amended, filings$OBJECTID)
winner <- filings$OBJECTID[o][nrow(filings)]
winner
```

After resolution, the person-year grain is `OBJECTID + TABLE_ID` (the source
person-line). This keeps two genuinely different people who share a name in the
same filing as **distinct** records, and makes `PERSON_YEAR_ID` map 1:1 to panel
rows — collision-free by construction.

## `EMP_ID`: the cross-year person id

`EMP_ID` is stamped by `link_panel()` after the accepted links close into person
clusters. It is a **content hash of the cluster's sorted members**, so it is
stable across runs as long as the membership is. The lower-level hashing is
exported:

```{r emp}
# One canonical string per record, fed to the hasher:
df <- data.frame(ein = c("11", "11"), taxyr = c(2019, 2020),
                 name = c("Jane Roe", "Jane Roe"),
                 title.standard = c("CEO", "CEO"))
ids <- build_id_string(df)
ids

# Deterministic EMP-style hash of a vector of keys (same input -> same id):
create_emp_ids(c("cluster-A", "cluster-B", "cluster-A"))
```

## How the two relate

Within a person's `EMP_ID` cluster, each row still carries its own
`PERSON_YEAR_ID`. So `EMP_ID` groups the rows; `PERSON_YEAR_ID` ties each grouped
row back to its exact source filing. A longitudinal panel keyed on `EMP_ID` can
always drill back to the raw Part VII line via `PERSON_YEAR_ID`.

## Future work: `CONTENT_ID`

`PERSON_YEAR_ID` depends on `TABLE_ID`, which is parser-derived and has broken on
pathological XML before. A parser-independent `CONTENT_ID` — a hash of normalized
native content fields — could be stored alongside as a recovery bridge if
`TABLE_ID` numbering ever changes globally. It is documented as future work in
`?person_year_id`; because a parser change bad enough to break `TABLE_ID` usually
corrupts that record's content too, `CONTENT_ID` mainly insures against a *global*
renumbering of otherwise-clean rows. Persisting the raw `OBJECTID`/`TABLE_ID`
columns keeps that bridge buildable after the fact.

## See also

- [Getting started](getting-started.html) for where these ids appear in output.
- `person_year_id()`, `create_emp_ids()`, `build_id_string()` in the reference.
```
