---
title: "Reviewing and QA-ing links"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reviewing and QA-ing links}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(synthid)
```

Automated linkage should never be the last word on a longitudinal panel. This
article covers the quality-control layer: how to surface the links most worth a
human's attention, and how to route upstream **name-parser defects** back to
`peopleparser` instead of silently absorbing them.

There are two distinct jobs, and two functions:

- **`flag_links()`** — flags *linkage* decisions that deserve review.
- **`parse_fail_log()`** — flags *parser* defects that corrupted the inputs.

## Input contract

Both operate on a linked panel (the output of `link_panel()`), and both need the
raw name to compare against the parsed components. The relevant columns:

| role | default column |
|---|---|
| person id | `EMP_ID` |
| organization | `ein`, `org.name` |
| tax year | `taxyr` |
| raw name | `name.raw` |
| parsed parts | `first_name`, `last_name` |

`name.raw` is the literal source name; carrying it through the pipeline (see the
[Getting started](getting-started.html) article) is what makes this review layer
possible.

## Flagging links for review

`flag_links()` scans each linked cluster for the failure modes that survive
automated scoring and tags them with a category:

```{r flag, eval = FALSE}
linked <- link_panel(panel)
flags  <- flag_links(linked)
table(flags$category)
```

The categories name *why* a link (or non-link) is suspect — for example:

| category | what it catches |
|---|---|
| `surname_compound` | compound surnames matched on one shared token |
| `surname_credential` | a credential (`MD`, `CPA`) fused into the surname |
| `parser_honorific_glue` | an honorific glued onto the name by the parser |
| `nickname_or_spelling` | first names differing by a nickname or spelling |

Turn the flagged set into a ready-to-work queue (sorted, deduplicated, one row per
item to review) and export it:

```{r queue, eval = FALSE}
q <- link_review_queue(flags)
utils::write.csv(q, "link_review_queue.csv", row.names = FALSE)
```

## Routing parser defects upstream

Some problems are not linkage errors at all — they are defects in the parsed name
that `synthid` merely *detects*. `parse_fail_log()` isolates these so they can be
fixed at the source (`peopleparser`) rather than patched here:

```{r parsefail, eval = FALSE}
log <- parse_fail_log(linked, path = "peopleparser_parse_fails.csv")
table(log$defect_type)
```

The defect classes correspond to concrete parser failure modes:

| defect_type | meaning |
|---|---|
| `honorific_glue` | an honorific (`REV`, `HON`) stuck to a given name |
| `credential_in_surname` | a post-nominal credential absorbed into the surname |
| `surname_letter_glue` | a stray initial glued to the surname |

## The token vocabularies

The flaggers are driven by small, inspectable token sets — the honorifics and
credentials they look for. These are exported so you can see exactly what is
matched (and extend them):

```{r tokens}
head(honorific_tokens(), 20)
head(credential_tokens(), 20)
```

Pass extended vectors via the `hon =` / `cred =` arguments of `flag_links()` to
tune sensitivity.

## A workflow

1. `link_panel()` to produce the linked panel.
2. `flag_links()` → `link_review_queue()` → hand the CSV to a reviewer.
3. `parse_fail_log()` → hand its CSV to the `peopleparser` maintainer.
4. Re-run once upstream fixes land; the deterministic ids mean unaffected records
   keep their `EMP_ID` / `PERSON_YEAR_ID`.

## See also

- `flag_links()`, `link_review_queue()`, `parse_fail_log()`,
  `honorific_tokens()`, `credential_tokens()` in the reference.
- [The two identifiers](identifiers.html) for why `name.raw` is retained.
```
