---
title: "How the matching works"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How the matching works}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(synthid)
```

Linkage comes down to one repeated decision: *are these two records the same
person?* This article opens up how `synthid` answers it — the comparators, the
score, and the learned model — so the decisions are auditable rather than a black
box.

## The score in one line

Each candidate pair gets an additive score. Every field contributes

```
weight x (2 * similarity - 1)
```

so a **similarity of 1** adds `+weight`, a **similarity of 0** subtracts
`weight`, and a **missing field is neutral** (it drops out rather than counting
as agreement). A pair links when its total clears the `threshold`. The interesting
engineering is in (a) how each field's *similarity* is computed and (b) how the
surname *weight* is adjusted per organization.

## The name comparators

Plain string distance is not enough for names. Two comparators do the heavy
lifting.

### Surname: robust to tokens and separators

`compare_two_names()` takes the max of a whole-string match (robust to separator
noise) and a token-coverage match divided by the *smaller* token count (robust to
a dropped or reordered token) — then rescales so that unrelated names land near 0:

```{r surname}
compare_two_names("MCLANE", "MC-LANE")               # separator noise -> ~1
compare_two_names("ANDREWS-MCLANE", "MCLANE")        # dropped token   -> high
compare_two_names("WEINER-COHEN", "COHEN-GANTSOUDES") # share ONE token -> low
```

That last case is the point: two different multi-token surnames that happen to
share one token must **not** masquerade as the same person.

### First name: nicknames and sound-alikes, but not siblings

`compare_first_names()` treats a *direct* nickname as near-agreement, an initial
vs a full name as weak, and a phonetic sound-alike as penalty-free — while
refusing to equate sibling diminutives:

```{r firstname}
compare_first_names(c("BOB", "JON",     "LISA"),
                    c("ROBERT", "JONATHAN", "BETH"))
# BOB~ROBERT and JON~JONATHAN score high; LISA~BETH does NOT.
```

### Why rescaling matters

Plain Jaro–Winkler rarely drops below ~0.5, so a clear mismatch would read as
"half a match." Both name similarities are rescaled so unrelated names actually
*penalise* — otherwise a nickname plus gender agreement could override a genuinely
different surname.

## The surname-frequency adjustment (the family-board problem)

Boards are often family-dominated. Agreeing on `SMITH` inside a board that is
half Smiths tells you almost nothing. `surname_weight()` scales the surname's
contribution *per organization* by how rare it is inside that org:

```{r surnamewt}
board <- data.frame(
  ein = "100",
  name = c("JOHN SMITH", "JANE SMITH", "ANN JONES"),
  last_name = c("SMITH", "SMITH", "JONES")
)
surname_weight(board, org_id = "ein", name = "name", last_name = "last_name")
```

The adjustment is deliberately **asymmetric**: an *agreement* is weighted by the
commoner side (so a shared common surname carries little information and the first
name / suffix / gender do the discriminating), while a *disagreement* is weighted
by the rarer side (so a genuine surname mismatch keeps its full penalty and cannot
be diluted away). `population_surname_weight()` provides a corpus-level prior for
the cross-organization setting.

## The default weights

The base field weights are inspectable and overridable:

```{r weights}
default_weights()
```

Pass a modified copy to `link_panel(panel, weights = ...)` to reweight fields.

## Scoring a panel

`candidate_scores()` exposes the per-pair scores directly — useful for choosing a
threshold or debugging a specific pair:

```{r scores}
rec <- function(ein, yr, first, last, suffix, gender) data.frame(
  ein = ein, taxyr = yr, name = trimws(paste(first, last, suffix)),
  first_name = first, middle_name = NA_character_, last_name = last,
  suffix = suffix, gender = gender, stringsAsFactors = FALSE)
panel <- rbind(
  rec("100", 2019, "JOHN", "SMITH", "SR", "M"),
  rec("100", 2019, "ANN",  "JONES", "",   "F"),
  rec("100", 2020, "JOHN", "SMITH", "SR", "M"),
  rec("100", 2020, "ANN",  "JONES", "",   "F"),
  rec("100", 2021, "JOHN", "SMITH", "SR", "M"),
  rec("200", 2019, "MARY", "REED",  "",   "F"),
  rec("200", 2020, "MARY", "REED",  "",   "F")
)
cs <- candidate_scores(panel)
cs[, c("org", "yr_x", "yr_y", "first_name_x", "first_name_y",
       "last_name_x", "last_name_y", "score")]
```

The true same-person pairs (JOHN SMITH across years, MARY REED across years) score
well above the cross-person pairs, which the rescaled name comparators push
sharply negative.

## A learned check: unsupervised EM

The hand weights are validated by fitting an **unsupervised** Fellegi–Sunter
latent-class model to the comparison vectors — no labels — and reading off the
learned agreement/disagreement weights:

```{r em}
cmp <- candidate_comparisons(panel)
em  <- fit_match_model(cmp)     # method = "em"
fs_weights(em)
```

On the real development panel the EM and hand-weighted methods agree on **99.85%**
of accept/reject decisions, and the EM posterior is sharply bimodal — which is
what makes `EMP_LINK_CONF` (emitted by `link_panel(method = "em")`) a meaningful
confidence signal. `predict_match()` returns the calibrated per-pair probability.

## Choosing a threshold

The default `threshold = 7` was tuned against a labeled slice: pair-level
precision 0.98 / recall 1.00 overall, and precision 0.985 / recall 0.955 on a
family-board slice, with post-one-to-one precision ~0.999. Raise toward 8–8.5 for
maximum precision (at some cost to family-board recall); lower for higher recall.

## See also

- [Getting started](getting-started.html) — the end-to-end walk-through.
- `default_weights()`, `fs_weights()`, `candidate_comparisons()`,
  `fit_match_model()`, `predict_match()` in the reference.
```
