---
title: "Scaling to the full panel"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scaling to the full panel}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(synthid)
```

`link_panel()` runs in memory and is perfect for one organization's history or a
development slice. A full multi-year national panel is millions of records, so the
`dev/` scripts wrap `synthid` in a batched, parallel workflow. This article
explains the model; the scripts themselves are the runnable reference.

## The one fact that makes it embarrassingly parallel

**The independent unit of work is one `EIN`.** Blocking, cluster closure, and the
one-record-per-organization-year invariant are all *within-EIN*. There is no
global state to precompute — even the surname-rarity weight is computed *per-EIN*
inside `synthid` — so batches of whole EINs are fully independent.

This yields one hard correctness rule:

> Every `EIN` must land **wholly** in one batch. Splitting an EIN across batches
> would silently corrupt its linkage.

## The four stages

```sh
# A/B. Preprocess ONE year -> one Parquet file whose columns are exactly what
#      link_panel() consumes. Runs classify_titles + parse_names, resolves to the
#      authoritative filing per org-year, and stamps PERSON_YEAR_ID at ingest.
Rscript dev/01_preprocess_year.R 2019
Rscript dev/01_preprocess_year.R 2020
Rscript dev/01_preprocess_year.R 2021

# C. Stack every preprocessed year into one persistent DuckDB file and tile the
#    DISTINCT eins into batches (so each EIN lands wholly in one batch).
Rscript dev/02_build_duckdb.R 5000     # eins per batch

# D. Link the batches, in parallel ACROSS batches; one Parquet part per batch.
Rscript dev/03_run_batched.R 8         # n workers
```

Stage A/B is a rebuildable cache — run it once per tax year, never inside the
linkage loop. Stages C and D rebuild cheaply from those Parquet files.

## Why this layout

- **DuckDB as the spine.** Years are stacked into one on-disk table; each worker
  opens its *own read-only* connection (concurrent readers are safe) and pulls
  just its batch with a single `JOIN` on the batch assignment.
- **Parallelize across batches, not inside reclin2.** For typical org sizes,
  `reclin2`'s own socket parallelism costs more than it saves. The scripts keep
  every numeric library single-threaded inside each worker (`OMP_NUM_THREADS=1`,
  `setDTthreads(1)`) and get the win from one core per batch, via `furrr`
  `multisession` (Windows-safe, no forking).
- **Nothing flows back through the main process** except a tiny per-batch summary;
  each batch writes its own `part-NNNNN.parquet`.

## `PERSON_YEAR_ID` and filing resolution happen at ingest

Stage A/B is where each record gets its stable back-link to the source row, and
where multiple filings for one org-year are resolved to the authoritative return.
Both are described in [The two identifiers](identifiers.html). Doing this once, at
preprocess, means the batched linkage never has to reason about amended returns.

## The memoized comparator speedup

Within a run, the same name pair is compared many times. `synthid` memoizes the
expensive name comparators by default; on the development panel this is a ~15x
linkage speedup. Crucially it is **byte-identical** to the unmemoized path — it is
a pure cache, toggleable for A/B testing via an option:

```{r memoize}
panel <- rbind(
  data.frame(ein = "100", taxyr = 2019, name = "JOHN SMITH", first_name = "JOHN",
             middle_name = NA, last_name = "SMITH", suffix = "SR", gender = "M"),
  data.frame(ein = "100", taxyr = 2020, name = "JOHN SMITH", first_name = "JOHN",
             middle_name = NA, last_name = "SMITH", suffix = "SR", gender = "M"),
  data.frame(ein = "100", taxyr = 2021, name = "JOHN SMITH", first_name = "JOHN",
             middle_name = NA, last_name = "SMITH", suffix = "SR", gender = "M"),
  stringsAsFactors = FALSE
)

o1 <- options(synthid.memoize_comparators = FALSE)
base <- link_panel(panel)$EMP_ID
options(o1)

o2 <- options(synthid.memoize_comparators = TRUE)   # the default
memo <- link_panel(panel)$EMP_ID
options(o2)

identical(base, memo)     # the cache never changes the answer
```

A second option, `synthid.first_name_comparator`, falls the first-name comparator
back to plain Jaro–Winkler if you ever need to isolate its contribution.

## Combining the output

The per-batch Parquet parts in `data/linked/` share one schema and can be read
back as a single table:

```sh
# e.g. with DuckDB:
#   SELECT * FROM read_parquet('data/linked/part-*.parquet')
```

Because `EMP_ID` and `PERSON_YEAR_ID` are deterministic, re-running a single batch
(after an upstream fix, say) leaves every other batch's ids untouched.

## See also

- `dev/01_preprocess_year.R`, `dev/02_build_duckdb.R`, `dev/03_run_batched.R`.
- [The two identifiers](identifiers.html) and
  [How the matching works](matching-methodology.html).
```
