---
title: "Title roles: BOARD / OFFICER / STAFF / OTHER"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Title roles: BOARD / OFFICER / STAFF / OTHER}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(synthid)
```

The `title.standard` vocabulary from `titleclassifier` has ~250 values — too fine
to reason about the *composition* of a match. `classify_title_role()` collapses it
into four coarse roles used by cross-organization linkage and interlock studies.

## The four roles

```{r classify}
classify_title_role(c(
  "BOARD PRESIDENT", "BOARD SECRETARY", "TREASURER",   # governance -> BOARD
  "CHIEF FINANCIAL OFFICER", "EXECUTIVE DIRECTOR",      # management -> OFFICER
  "VICE PRESIDENT",                                     # governance -> BOARD
  "VICE PRESIDENT OF FINANCE",                          # functional -> STAFF
  "DIRECTOR OF MARKETING",                              # functional -> STAFF
  "CHAPLAIN"                                            # unmatched  -> OTHER
))
```

- **`BOARD`** — the governance body and its elected officers. Per the interlock
  use case, the governance triad (president / treasurer / secretary) is `BOARD`,
  **not** `OFFICER`.
- **`OFFICER`** — paid executive leadership: the C-suite, executive director,
  general manager, general counsel, comptroller.
- **`STAFF`** — other paid employees: functional directors and vice presidents,
  managers, coordinators, clerks.
- **`OTHER`** — anything unmatched (e.g. `CHAPLAIN`, `HISTORIAN`), by design, so
  the classifier stays safe on the long tail.

### The distinction that does the work

A governance title carrying a **functional** qualifier is a paid post, not a
governance seat. Any title containing `" OF "` is excluded from `BOARD`:

```{r ofrule}
classify_title_role(c("VICE PRESIDENT", "VICE PRESIDENT OF FINANCE"))
```

By default the result is an ordered factor (`levels = TRUE`) so roles sort
`BOARD < OFFICER < STAFF < OTHER`; pass `levels = FALSE` for a plain character
vector.

## Inspecting the rules

The rule set is exported and inspectable — the classifier is not a black box:

```{r rules}
str(title_role_rules(), max.level = 1)
```

## Auditing coverage and pinning exceptions

Because unmatched titles fall to `OTHER` on purpose, you should audit what lands
there for your corpus. `title_role_coverage()` returns one row per distinct title
with its frequency and assigned role:

```{r coverage}
titles <- c(rep("BOARD PRESIDENT", 5), rep("EXECUTIVE DIRECTOR", 3),
            rep("DIRECTOR OF NURSING", 2), "SERGEANT AT ARMS", "CHAPLAIN")
cov <- title_role_coverage(titles)
cov

# The tail worth reviewing:
subset(cov, role == "OTHER")
```

When you find a title in `OTHER` that belongs elsewhere, pin it with `overrides`
instead of editing rules — the override wins:

```{r overrides}
classify_title_role(c("SERGEANT AT ARMS", "CHAPLAIN"),
                    overrides = c("SERGEANT AT ARMS" = "BOARD"))
```

## Where roles are used

Role assignment drives the edge filters in cross-organization linkage —
`board_only_edges()` keeps board-to-board interlocks, `non_board_edges()` the
rest. See [Cross-organization linkage](cross-org.html).

## See also

- `classify_title_role()`, `title_role_rules()`, `title_role_coverage()` in the
  reference.
```
